Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-24 model.tex
fixing a table entry
update table, fit equations
airplane edits

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as
an estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.
Some example plots of data from
Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following
this procedure, are shown in Figure~\ref{fig:embedding}.
% \begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colipc3d}
%    \caption{\col }
%    \label{fig:colEmbedding}
%  \end{subfigure}%  \\
%
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gcc3dipc}
%    \caption{\gcc}
%    \label{fig:gccEmbedding}
%  \end{subfigure}
%  \\
%  \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svd53dipc2}
%    \caption{\svdfive}
%    \label{fig:svdfiveEmbedding}
%  \end{subfigure}%
%%
%
%
%     %\includegraphics[width=\textwidth]{colipc3d}
%     \caption{3D projections of delay-coordinate embeddings of the
%       traces from (a) Figure~\ref{fig:col-ts} (b)
%       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
%       Figure~\ref{fig:svd-ts-colored}.}
% \label{fig:embedding}
% \end{figure}
 \begin{figure}[!ht]
    \subfloat[\col\label{fig:colEmbedding}]{%
      \includegraphics[width=\columnwidth]{figs/colipc3d}
    }
    \hfill
    \subfloat[\gcc\label{fig:gccEmbedding}]{%
      \includegraphics[width=\columnwidth]{figs/gcc3dipc}
    }
        \hfill
    \subfloat[\svdfive \label{fig:svdfiveEmbedding}]{%
      \includegraphics[width=\columnwidth]{figs/svd53dipc2}
    }
    \caption{3D projections of delay-coordinate embeddings of the
       traces from (a) Figure~\ref{fig:col-ts} (b)
       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
       Figure~\ref{fig:svd-ts-colored}.}
    \label{fig:embedding}
  \end{figure}
%% can cut for space if need be:
%The coordinates of each point in these plots are differently delayed
%elements of the instructions per cycle time series $X_{i,obs}$: that
%is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
%$X_{i+2\tau,obs}$ on the third, and so on.
%

@@

Since LMA does not rest on an assumption of linearity (as ARIMA does),
it can handle both linear and nonlinear processes.  If the underlying
generating process is nondeterministic, however, it can perform
poorly.  In Figure~\ref{fig:embedding}b, for instance, little
structure is visible, so LMA may not work well.  More structure
appears to be present in~Figure~\ref{fig:embedding}c, but this
reconstruction also appears to contain some noise.  The question as to
how much structure is present in a reconstruction---and how much of
that structure can be captured and used by LMA---is apropos of the
central question treated in this paper.  It may be that \gcc has some
redundancy that LMA cannot exploit, or that the structure in \svdfive
is effectively obfuscated, from the standpoint of the LMA method, by
noise.  Without any knowledge of the generating process, answers to
these questions can only be derived from the data, with all of the
attendant problems (noise, sampling issues, etc.)  By quantifying the
balance between redundancy (that is, predictive structure) and entropy
for these real-valued time series, as shown in
Section~\ref{sec:results}, we can begin to answer these questions
in an effective and practical manner.

@@

Figure~\ref{fig:forecast-example} shows example forecasts made using
all four methods for the \col, \gcc, and \svdfive time series.
%\begin{figure*}[htbp]
%  \centering
%
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colRWForecast}
%    \caption{\col\\ random walk }
%    \label{fig:colRW}
%  \end{subfigure}%
%   \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
%    \caption{\gcc\\ random walk }
%    \label{fig:gccRW}
%  \end{subfigure}%
%     \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
%    \caption{\svdfive\\ random walk}
%    \label{fig:svd5RW}
%  \end{subfigure}%
%  \\
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
%    \caption{\col\\ na\"ive }
%    \label{fig:colMEAN}
%  \end{subfigure}%
%   \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
%    \caption{\gcc\\ na\"ive }
%    \label{fig:gccMEAN}
%  \end{subfigure}%
%     \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
%    \caption{\svdfive\\ na\"ive }
%    \label{fig:svd5MEAN}
%  \end{subfigure}%
%  \\
%    \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
%    \caption{\col \\ \arima}
%    \label{fig:colARIMA}
%  \end{subfigure}
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
%    \caption{\gcc \\ \arima }
%    \label{fig:gccARIMA}
%  \end{subfigure}%
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
%    \caption{\svdfive \\ \arima }
%    \label{fig:svd5ARIMA}
%  \end{subfigure}%
%  \\
%
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
%    \caption{\col \\ LMA}
%    \label{fig:colLMA}
%  \end{subfigure}
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
%    \caption{\gcc \\ LMA}
%    \label{fig:gccLMA}
%  \end{subfigure}
%        \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
%    \caption{\svdfive \\ LMA}
%    \label{fig:svd5LMA}
%  \end{subfigure}
%  %\begin{subfigure}{0.5\textwidth}
%  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
%   \caption{Predicted ($p_j$) versus true values ($c_j$) for
%     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
%% On this type of plot, a perfect prediction lies exactly on the
%% diagonal.
%%
%%
%}
%\label{fig:forecast-example}
%  %\end{subfigure}

@@

\newcolumntype{C}{>{\centering\arraybackslash} m{0.6\columnwidth} }
\begin{figure*}
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{m{0.1\columnwidth}|CCC}
    & {\large \col} & {\large \gcc} & {\large \svdfive} \\
    & & & \\
            \hline \\
    \rotatebox{90}{\large{random walk}} &
    \includegraphics[width=0.6\columnwidth]{figs/colRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveRWForecast} \\
    \rotatebox{90}{\large \naive} &
    \includegraphics[width=0.6\columnwidth]{figs/colMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
    \rotatebox{90}{\large \arima} &
    \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveARIMAForecast} \\
    \rotatebox{90}{\large LMA} &
    \includegraphics[width=0.6\columnwidth]{figs/colLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveLMAForecast}
  \end{tabular}
  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts of \col,
     \gcc, and \svdfive and all four prediction strategies.
  }
  \label{fig:forecast-example}
\end{figure*}
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while \arima
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of some of the plots in Figure~\ref{fig:forecast-example}, random walk and \arima on \col for example,  are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For \arima, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the \arima $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error (MASE) between the true and predicted signals:
%
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
%
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  MASE is a normalized measure: the scaling
term in the denominator
%
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$
%
is the average in-sample forecast error for a random-walk prediction
over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$ means that the
prediction error in question was, on the average, smaller than the
error of a random-walk forecast on the same data.  Analogously,
$MASE>1$ means that the corresponding prediction method did
\emph{worse}, on average, than the random-walk method.  We chose this
error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales.

MASE scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
\caption{Mean
absolute scaled error (MASE) scores and weighted permutation entropies for all eight
   processes studied in this paper}
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } ccccc}
  \hline\hline Signal & LMA MASE  & \arima MASE & na\"{i}ve MASE  & Weighted Permutation Entropy \\
  \hline
 \col           & $ 0.050 \pm0.002  $ & $0.599  \pm 0.211 $ & $0.571\pm0.002$&  $0.513 \pm 0.003$ \\

\gcc           & $ 1.530\pm 0.021$ & $1.837 \pm0.016 $ & $0.951 \pm 0.001$ & $0.943 \pm 0.001$ \\

\svdone     & $ 0.827\pm 0.076$ & $ 0.714\pm 0.075 $ & $2.676\pm4.328$&  $0.957 \pm 0.016$ \\

 \svdtwo    & $1.279 \pm0.020 $ & $2.163 \pm0.027 $ &  $3.054\pm0.040$ &   $0.846 \pm0.004$ \\

\svdthree     & $0.619 \pm0.021 $ & $0.713 \pm 0.010 $ & $31.386\pm 0.282$ &  $0.716 \pm 0.006$ \\
 \svdfour     & $ 0.779\pm0.036 $ & $0.979 \pm0.032 $ & $2.661\pm0.074$ & $0.825 \pm 0.008$ \\

\svdfive     & $ 0.718\pm 0.048 $ & $2.370  \pm 0.051 $ & $20.870 \pm 0.192$&  $0.678 \pm 0.007$ \\
 \svdsix     & $ 0.739\pm 0.068 $ & $ 1.438\pm 0.061$ & $2.197\pm0.083$&  $0.748 \pm 0.011$ \\
  \hline\hline
  \end{tabular*}
  \end{center}
 \label{default}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals was
$0.050$, for instance, while \arima scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while \arima only outperformed random walk
by a factor of 1.7.  This is in accordance with the visual appearance
of the corresponding images in Figure~\ref{fig:forecast-example}.
While the correspondence between MASE score and the visual appearance
of these kinds of plots is clear cut in this case, that is not always
so.  The plots of LMA predictions of \col and \svdfive both lie near
the diagonal, for instance, but the corresponding MASE scores were
very different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20
and 1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a
double-edged sword.  Here, it facilitates comparisons across signals
of different types and lengths, but its particular pathologies must be
taken into account, as discussed at more length in
Section~\ref{sec:results}.

>>>>

The first step in the delay-coordinate embedding process is to
estimate values for the two free parameters in the map: the delay
$\tau$ and the dimension $m$.  We follow standard procedures for this,
choosing the first minimum in the time-delayed mutual information as
an estimate of $\tau$~\cite{fraser-swinney} and using the
false-near(est)-neighbor(s) method of~\cite{KBA92} to estimate $m$.
Some example plots of data from
Figures~\ref{fig:col-ts}-\ref{fig:svd-ts-colored}, embedded following
this procedure, are shown in Figure~\ref{fig:embedding}.
% \begin{figure}
%   \centering
%\begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colipc3d}
%    \caption{\col }
%    \label{fig:colEmbedding}
%  \end{subfigure}%  \\
%
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gcc3dipc}
%    \caption{\gcc}
%    \label{fig:gccEmbedding}
%  \end{subfigure}
%  \\
%  \begin{subfigure}{\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svd53dipc2}
%    \caption{\svdfive}
%    \label{fig:svdfiveEmbedding}
%  \end{subfigure}%
%%
%
%
%     %\includegraphics[width=\textwidth]{colipc3d}
%     \caption{3D projections of delay-coordinate embeddings of the
%       traces from (a) Figure~\ref{fig:col-ts} (b)
%       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
%       Figure~\ref{fig:svd-ts-colored}.}
% \label{fig:embedding}
% \end{figure}
 \begin{figure}[!ht]
    \subfloat[\col\label{fig:colEmbedding}]{%
      \includegraphics[width=\columnwidth]{figs/colipc3d}
    }
    \hfill
    \subfloat[\gcc\label{fig:gccEmbedding}]{%
      \includegraphics[width=\columnwidth]{figs/gcc3dipc}
    }
        \hfill
    \subfloat[\svdfive \label{fig:svdfiveEmbedding}]{%
      \includegraphics[width=\columnwidth]{figs/svd53dipc2}
    }
    \caption{3D projections of delay-coordinate embeddings of the
       traces from (a) Figure~\ref{fig:col-ipc} (b)
       Figure~\ref{fig:gcc-ts} and (c) the fifth (green) segment of
       Figure~\ref{fig:svd-ts-colored}.}
    \label{fig:embedding}
  \end{figure}
%% can cut for space if need be:
%The coordinates of each point in these plots are differently delayed
%elements of the instructions per cycle time series $X_{i,obs}$: that
%is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
%$X_{i+2\tau,obs}$ on the third, and so on.
%

@@

Since LMA does not rest on an assumption of linearity (as ARIMA models
do), it can handle both linear and nonlinear processes.  If the
underlying generating process is nondeterministic, however, it can
perform poorly.  Figure~\ref{fig:embedding}b, for instance, appears to
contain little structure, so LMA may not work well on this signal.
More structure appears to be present in Figure~\ref{fig:embedding}c,
but this reconstruction also appears to contain some noise.  The
question as to how much structure is present in a reconstruction---and
how much of that structure can be captured and used by LMA---is
apropos of the central question treated in this paper.  It may be that
\gcc has some redundancy that LMA cannot exploit, or that the
structure in \svdfive is effectively obfuscated, from the standpoint
of the LMA method, by noise.  Without any knowledge of the generating
process, answers to these questions can only be derived from the data,
with all of the attendant problems (noise, sampling issues, etc.)  By
quantifying the balance between redundancy (that is, predictive
structure) and entropy for these real-valued time series, as shown in
Section~\ref{sec:results}, we can begin to answer these questions in
an effective and practical manner.

@@

Figure~\ref{fig:forecast-example} shows example forecasts made using
all four methods for the \col, \gcc, and \svdfive time series.
%\begin{figure*}[htbp]
%  \centering
%
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colRWForecast}
%    \caption{\col\\ random walk }
%    \label{fig:colRW}
%  \end{subfigure}%
%   \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccRWForecast}
%    \caption{\gcc\\ random walk }
%    \label{fig:gccRW}
%  \end{subfigure}%
%     \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveRWForecast}
%    \caption{\svdfive\\ random walk}
%    \label{fig:svd5RW}
%  \end{subfigure}%
%  \\
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colMeanForecast}
%    \caption{\col\\ na\"ive }
%    \label{fig:colMEAN}
%  \end{subfigure}%
%   \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccMeanForecast}
%    \caption{\gcc\\ na\"ive }
%    \label{fig:gccMEAN}
%  \end{subfigure}%
%     \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveMeanForecast}
%    \caption{\svdfive\\ na\"ive }
%    \label{fig:svd5MEAN}
%  \end{subfigure}%
%  \\
%    \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colARIMAForecast}
%    \caption{\col \\ \arima}
%    \label{fig:colARIMA}
%  \end{subfigure}
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccARIMAForecast}
%    \caption{\gcc \\ \arima }
%    \label{fig:gccARIMA}
%  \end{subfigure}%
%  \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveARIMAForecast}
%    \caption{\svdfive \\ \arima }
%    \label{fig:svd5ARIMA}
%  \end{subfigure}%
%  \\
%
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/colLMAForecast}
%    \caption{\col \\ LMA}
%    \label{fig:colLMA}
%  \end{subfigure}
%      \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/gccLMAForecast}
%    \caption{\gcc \\ LMA}
%    \label{fig:gccLMA}
%  \end{subfigure}
%        \begin{subfigure}{0.6\columnwidth}
%    \includegraphics[width=\columnwidth]{figs/svdfiveLMAForecast}
%    \caption{\svdfive \\ LMA}
%    \label{fig:svd5LMA}
%  \end{subfigure}
%  %\begin{subfigure}{0.5\textwidth}
%  %  \includegraphics[width=1.0\textwidth]{LMA_vs_ARIMA}
%   \caption{Predicted ($p_j$) versus true values ($c_j$) for
%     forecasts of \col, \gcc, and \svdfive and all four prediction strategies.
%% On this type of plot, a perfect prediction lies exactly on the
%% diagonal.
%%
%%
%}
%\label{fig:forecast-example}
%  %\end{subfigure}

@@

\newcolumntype{C}{>{\centering\arraybackslash} m{0.6\columnwidth} }
\begin{figure*}
  \renewcommand{\arraystretch}{1.25}
  \begin{tabular}{m{0.1\columnwidth}|CCC}
    & {\large \col} & {\large \gcc} & {\large \svdfive} \\
    & & & \\
            \hline \\
    \rotatebox{90}{\large{random walk}} &
    \includegraphics[width=0.6\columnwidth]{figs/colRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccRWForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveRWForecast} \\
    \rotatebox{90}{\large \naive} &
    \includegraphics[width=0.6\columnwidth]{figs/colMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccMeanForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveMeanForecast} \\
    \rotatebox{90}{\large \arima} &
    \includegraphics[width=0.6\columnwidth]{figs/colARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccARIMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveARIMAForecast} \\
    \rotatebox{90}{\large LMA} &
    \includegraphics[width=0.6\columnwidth]{figs/colLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/gccLMAForecast} &
    \includegraphics[width=0.6\columnwidth]{figs/svdfiveLMAForecast}
  \end{tabular}
  \caption{Predicted ($p_j$) versus true values ($c_j$) for forecasts of \col,
     \gcc, and \svdfive and all four prediction strategies.
  }
  \label{fig:forecast-example}
\end{figure*}
In these images, the vertical axis is the prediction $p_j$ and the
horizontal axis is the true continuation $c_j$.  On such a plot, a
perfect prediction would lie on the diagonal.  LMA, for instance,
generates a very accurate prediction of the \col data, while \arima
does not.  Horizontal lines result when a constant predictor (e.g.,
\naive) is used on a non-constant signal.  Point clouds reflect the
structure of the distribution of the errors---roughly normally
distributed, for example, in the case of LMA on \gcc.  Note that the
shapes of some of the plots in Figure~\ref{fig:forecast-example}, random walk and \arima on \col for example,  are
reminiscent of the projected embedding in Figure~\ref{fig:embedding}a.
For a random-walk predictor, a $p_j$ vs. $c_j$ plot is equivalent to a
two-dimensional embedding with $\tau=1$.  For \arima, the
correspondence is not quite as simple, since the $p_j$ values are
linear combinations of a number of past values of the $c_j$, but the
effect is largely the same\footnote{The temporal \emph{ordering} of
  the points on the \arima $p_j$ vs. $c_j$ plot does not match that of
  a projected embedding of the time series, however.}.

As a numerical measure of prediction accurary, we calculate the mean
absolute scaled error (MASE) between the true and predicted signals:
%
$$MASE = \sum_{j=n+1}^{k+n+1}\frac{|p_j-c_j|
}{\frac{k}{n-1}\sum^n_{i=2}|x_{i}-x_{i-1}|}$$
%
This error metric was introduced in \cite{MASE} as a ``generally
applicable measurement of forecast accuracy without the problems seen
in the other measurements."  MASE is a normalized measure: the scaling
term in the denominator
%
% of the MASE value
% $$\frac{1}{n-1}\sum^n_{i=2}|X_{i,obs}-X_{i-1,obs}|$$
%
is the average in-sample forecast error for a random-walk prediction
over the initial training signal $\{x_i\}^n_{i=1}$.  That is, $MASE<1$
means that the prediction error in question was, on the average,
smaller than the error of a random-walk forecast on the same data.
Analogously, $MASE>1$ means that the corresponding prediction method
did \emph{worse}, on average, than the random-walk method.  We chose
this error metric because it allows for fair comparison across varying
methods, prediction horizons, and signal scales---attributes that are
key to any broad study of predictability.

MASE scores for all 360 experiments are tabulated in the three
middle columns in Table~\ref{tab:error}.
\begin{table*}
\caption{Mean
absolute scaled error (MASE) scores and weighted permutation entropies for all eight
   processes studied in this paper}
  \begin{center}
  \begin{tabular*}{\textwidth}{@{\extracolsep{\fill} } cccccc}
  \hline\hline Signal & LMA MASE & \arima MASE & na\"{i}ve MASE & RW MASE & WPE \\
  \hline

  \col       & $0.050 \pm 0.002$ & $0.599 \pm 0.211$ & $0.571 \pm 0.002$  & $1.001 \pm 0.002$ & $0.513 \pm 0.003$ \\

  \gcc       & $1.530 \pm 0.021$ & $1.837 \pm 0.016$ & $1.797 \pm 0.010$  & $1.138 \pm 0.011$ & $0.943 \pm 0.001$ \\

  \svdone    & $0.827 \pm 0.076$ & $0.714 \pm 0.075$ & $2.676 \pm 4.328$  & $0.933 \pm 0.095$ & $0.957 \pm 0.016$ \\

  \svdtwo    & $1.279 \pm 0.020$ & $2.163 \pm 0.027$ & $3.054 \pm 0.040$  & $1.125 \pm 0.012$ & $0.846 \pm 0.004$ \\

  \svdthree  & $0.619 \pm 0.021$ & $0.713 \pm 0.010$ & $31.386 \pm 0.282$ & $0.707 \pm 0.009$ & $0.716 \pm 0.006$ \\

  \svdfour   & $0.779 \pm 0.036$ & $0.979 \pm 0.032$ & $2.661 \pm 0.074$  & $1.034 \pm 0.035$ & $0.825 \pm 0.008$ \\

  \svdfive   & $0.718 \pm 0.048$ & $2.370 \pm 0.051$ & $20.870 \pm 0.192$ & $1.001 \pm 0.047$ & $0.678 \pm 0.007$ \\

  \svdsix    & $0.739 \pm 0.068$ & $1.438 \pm 0.061$ & $2.197 \pm 0.083$  & $1.060 \pm 0.055$ & $0.748 \pm 0.011$ \\

  \hline\hline
  \end{tabular*}
  \end{center}
 \label{tab:error}
  \end{table*}%
Comparing these with the geometry of the plots in
Figure~\ref{fig:forecast-example}, one can see some obvious
correspondences.  The average LMA MASE score for the \col signals was
$0.050$, for instance, while \arima scored much worse ($0.599$).
That is, LMA performed roughly 20 times better on \col signals than a
random-walk predictor, while \arima only outperformed random walk
by a factor of 1.7.  This is in accordance with the visual appearance
of the corresponding images in Figure~\ref{fig:forecast-example}.
While the correspondence between MASE score and the visual appearance
of these kinds of plots is clear cut in this case, that is not always
so.  The plots of LMA predictions of \col and \svdfive both lie near
the diagonal, for instance, but the corresponding MASE scores were
very different: $0.050$ for \col and $0.718$ for \svdfive (resp., 20
and 1.4 times better than random-walk forecasts of the same signals).
This is a result of the normalization in the MASE score calculation.
Recall from Section~\ref{sec:simple} that the random-walk method
performs especially poorly on signals that oscillate rapidly.  \col
fits this description, so the random-walk error on this signal---the
denominator of the three MASE scores in the first row of the
Table---is pathologically high, which skews those scores down.
Random-walk prediction is very effective for the \svdfive signal, on
the other hand, so the denominator is small and the MASE scores are
skewed upwards.  Normalization is, as is often the case, a
double-edged sword.  Here, it facilitates comparisons across signals
of different types and lengths, but its particular pathologies must be
taken into account, as discussed at more length in
Section~\ref{sec:results}.

### 2014-07-24 results.tex
airplane edits

In this section, we offer an empirical validation of the two
conjectures offered in Section \ref{sec:intro}, namely:

\begin{enumerate}

\item That the weighted permutation entropy (WPE) of a noisy
  real-valued time series from an unknown system is correlated with
  prediction accuracy---i.e., that the predictable structure in an
  empirical time-series data set can be quantified by its WPE.

%\subitem Simply reminders not to be included]]WPE can quantify when
%a noisy real-valued time series is predictable.I am unsatisfied
%with predictable in this sentence, need a better word to say "better
%than random walk" or ``able to be forecast effectively" or ``has the
%structural capacity to transmit information in a way that the time
%series can be effectively forecast"

%\item The existence of complex structure in a noisy real-valued time
%series is quantifiable and this type of complexity is directly
%correlated with predictability

\item That the way information is generated and processed by an
  unknown system is correlated with the effectiveness of a given
  predictor on empirical time-series data from that system

%\subitem Simply reminders not to be included]]We will have shown
%that the existence whether linear or nonlinear is picked up on with
%WPE but this point gets at whether the prediction model can use the
%structure or not (linear can't use nonlinear structure). The right to
%left shifts in \col and some of the \svd regimes and the lack of
%shift in \gcc illustrate this nicely.

%\subitem Simply reminders not to be included]]This is the linear vs
%nonliear vs random we see with the right to left shifts with lower
%complexity time series.

@@

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the best prediction of each time series.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.
%
% (across all trials and all prediction methods)
%
The dashed curve is a least-squares log fit of all the points except
those from {\tt dgesdd$_1$}, which we have excluded for reasons
explained in the text.  The shaded region boundaries reflect the
standard deviation of the model in its parameter space.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the best-case prediction error for each trace from
these eight systems is roughly proportional to the log of the weighted
permutation entropy, which is consistent with our first conjecture.
The dashed line in the figure, a least-squares fit of $y = a \log(b x
+ 1) + c x^16$ to these errors\footnote{with the exception of {\tt dgesdd$_1$},
  for reasons described below} captures this relationship between WPE
and prediction error quite nicely.  The logarithmic shape of this
curve makes sense here because any signal with zero entropy should be
perfectly predictable, and so should ideally have a MASE value of
zero.  The shaded region shows the standard deviation of the model in
its parameter space: i.e., $y = a \log(b x + 1) + c x^16$ with $\pm$ one
standard deviation on each of the three parameters.

This is not a formal result.  The forecast methods and data sets used
here were chosen to span the space of standard prediction strategies
and the range of dynamical behaviors, but they do not cover those
spaces exhaustively.  Our goal here is an\emph{empirical} assessment
of the relationship between predictability and complexity, not formal
results about a ``best'' predictor for a given time series.  There may
be other methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the curve in this plot strongly suggests a pattern of
(inverse) correlation between the underlying predictability of a time
series and its WPE.  The rest of this section describes our results in
more detail---including the measures taken to assure meaningful
comparisons across methods, trials, and programs---and elaborates on
the meaning of the dashed curve in the figure.

>>>>

In this section, we offer an empirical validation of the two findings
introduced in Section \ref{sec:intro}, namely:

\begin{enumerate}
% \item The complexity of a noisy real-valued time series can be
%   effectively quantifiable by computing its permutation entropy.

\item The weighted permutation entropy (WPE) of a noisy real-valued
  time series from an unknown system is correlated with prediction
  accuracy---i.e., the predictable structure in an empirical
  time-series data set can be quantified by its WPE.

\item The relationship between WPE and MASE is a useful empirical
  heuristic for identifying mismatches between prediction models and
  time-series data---i.e., there is structure in the data that the
  model is unable to exploit.

@@

Figure~\ref{fig:wpe_vs_mase_best} plots the WPE values versus the
corresponding MASE values of the best prediction of each of the
\alert{[[108?]]} time series in this study
\begin{figure}[htbp]
  \centering
  \includegraphics[width=\columnwidth]{figs/new_prediction_vs_entropy}
  \caption{Weighted permutation entropy vs. mean absolute scaled error
    (MASE) of the best prediction of each time series.
%
% (across all trials and all prediction methods)
%
The dashed curve is a least-squares \alert{[[log]]} fit of all the
points except those from {\tt dgesdd$_1$}, which we have excluded for
reasons explained in the text.  The shaded region boundaries reflect
the standard deviation of the model in its parameter space.}
  \label{fig:wpe_vs_mase_best}
\end{figure}
As the plot shows, the best-case prediction error for each trace from
these eight systems is roughly proportional to the \alert{[[log]]} of
the weighted permutation entropy, which is consistent with our first
conjecture.  \alert{We need to explain and justify this baroque curve
  verrry carefully.  Occam's razor it is not.}  The dashed line in the
figure, a least-squares fit of $y = a \log(b x + 1) + c x^{16}$ to
these errors\footnote{with the exception of {\tt dgesdd$_1$}, for
  reasons described below} captures this relationship between WPE and
prediction error quite nicely.  The logarithmic shape of this curve
makes sense here because any signal with zero entropy should be
perfectly predictable, and so should ideally have a MASE value of
zero.  The shaded region shows the standard deviation of the model in
its parameter space: i.e., $y = a \log(b x + 1) + c x^{16}$ with $\pm$
one standard deviation on each of the three parameters.

This is not a formal result.  The forecast methods and data sets used
here were chosen to span the space of standard prediction strategies
and the range of dynamical behaviors, but they do not cover those
spaces exhaustively.  Our goal here is an \emph{empirical} assessment
of the relationship between predictability and complexity, not formal
results about a ``best'' predictor for a given time series.  There may
be other methods that produce lower MASE values than those in
Figure~\ref{fig:wpe_vs_mase_best}, but the sparseness of the points
above and below the curve in this plot strongly suggests a pattern of
(inverse) correlation between the underlying predictability of a time
series and its WPE.  The rest of this section describes our results in
more detail---including the measures taken to assure meaningful
comparisons across methods, trials, and programs---and elaborates on
the meaning of the dashed curve in the figure.
